---
title: "LFPR and Employment Rates by Sex and Education: Workers in Their 20s"
format: html
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(haven)
library(knitr)
library(scales)

data_dir <- file.path(getwd(), "..", "data")

cps_file <- list.files(data_dir, pattern = "^cps_.*\\.dta$", full.names = TRUE)
stopifnot("Expected exactly one CPS .dta file in ../data/" = length(cps_file) == 1)

df <- read_dta(cps_file, col_select = c(
  "year", "month", "age", "sex", "empstat", "classwkr", "educ", "schlcoll", "wtfinl"
))
names(df) <- toupper(names(df))
```

```{r sample}
# Sample restrictions: drop armed forces, keep ages 20-29
df <- df |>
  filter(CLASSWKR != 26) |>
  filter(AGE >= 20, AGE <= 29)

# Variable construction
df <- df |>
  mutate(
    female = factor(ifelse(SEX == 2, "Women", "Men"), levels = c("Men", "Women")),
    college = ifelse(EDUC >= 111, "College (BA+)", "Non-College"),
    employed = EMPSTAT %in% c(10, 12),
    unemployed = EMPSTAT %in% c(20, 21, 22),
    nilf = EMPSTAT %in% c(32, 34, 36),
    in_lf = employed | unemployed,
    in_school = SCHLCOLL %in% c(1, 2, 3, 4),
    year_label = factor(YEAR)
  )
```

```{r collapse}
# Step 1: weighted monthly totals by sex x education x year x month
monthly <- df |>
  group_by(YEAR, MONTH, female, college) |>
  summarise(
    pop = sum(WTFINL[employed | unemployed | nilf]),
    emp = sum(WTFINL[employed]),
    lf  = sum(WTFINL[in_lf]),
    .groups = "drop"
  ) |>
  mutate(
    er   = 100 * emp / pop,
    lfpr = 100 * lf / pop
  )

# Step 2: average across months within each year (removes seasonality)
annual <- monthly |>
  group_by(YEAR, female, college) |>
  summarise(
    er   = mean(er),
    lfpr = mean(lfpr),
    .groups = "drop"
  ) |>
  mutate(year_label = factor(YEAR))
```

## Labor Force Participation Rate

LFPR = (employed + unemployed) / civilian noninstitutional population, ages 20-29. Weighted by WTFINL, averaged across all months within each year. Excludes armed forces.

```{r table-lfpr}
annual |>
  select(YEAR, female, college, lfpr) |>
  pivot_wider(names_from = YEAR, values_from = lfpr) |>
  arrange(college, female) |>
  kable(digits = 1, col.names = c("Sex", "Education", "1990", "2019", "2024"),
        caption = "Labor Force Participation Rate (%), Ages 20-29")
```

```{r plot-lfpr, fig.width=8, fig.height=5}
ggplot(annual, aes(x = year_label, y = lfpr, fill = female)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%.1f", lfpr)),
            position = position_dodge(width = 0.7), vjust = -0.4, size = 3) +
  facet_wrap(~college) +
  scale_fill_manual(values = c("Men" = "#348ABD", "Women" = "#E24A33")) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title = "Labor Force Participation Rate, Ages 20-29",
    subtitle = "Annual average across months. Excludes armed forces. Source: IPUMS CPS.",
    y = "LFPR (%)", x = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

## Employment Rate

Employment rate = employed / civilian noninstitutional population, ages 20-29.

```{r table-er}
annual |>
  select(YEAR, female, college, er) |>
  pivot_wider(names_from = YEAR, values_from = er) |>
  arrange(college, female) |>
  kable(digits = 1, col.names = c("Sex", "Education", "1990", "2019", "2024"),
        caption = "Employment Rate (%), Ages 20-29")
```

```{r plot-er, fig.width=8, fig.height=5}
ggplot(annual, aes(x = year_label, y = er, fill = female)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%.1f", er)),
            position = position_dodge(width = 0.7), vjust = -0.4, size = 3) +
  facet_wrap(~college) +
  scale_fill_manual(values = c("Men" = "#348ABD", "Women" = "#E24A33")) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title = "Employment Rate, Ages 20-29",
    subtitle = "Annual average across months. Excludes armed forces. Source: IPUMS CPS.",
    y = "Employment Rate (%)", x = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

## Gender Gap (Women - Men)

```{r gap}
gap <- annual |>
  select(YEAR, female, college, er, lfpr) |>
  pivot_wider(names_from = female, values_from = c(er, lfpr)) |>
  mutate(
    gap_er   = er_Women - er_Men,
    gap_lfpr = lfpr_Women - lfpr_Men
  )

gap |>
  select(YEAR, college, gap_lfpr, gap_er) |>
  kable(digits = 1,
        col.names = c("Year", "Education", "LFPR Gap (pp)", "ER Gap (pp)"),
        caption = "Gender Gap (Women - Men), Ages 20-29")
```

```{r plot-gap, fig.width=8, fig.height=5}
gap_long <- gap |>
  select(YEAR, college, gap_lfpr, gap_er) |>
  pivot_longer(cols = c(gap_lfpr, gap_er),
               names_to = "measure", values_to = "gap") |>
  mutate(
    measure = recode(measure,
                     gap_lfpr = "LFPR",
                     gap_er = "Employment Rate"),
    year_label = factor(YEAR)
  )

ggplot(gap_long, aes(x = year_label, y = gap, fill = college)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%+.1f", gap)),
            position = position_dodge(width = 0.7), vjust = -0.4, size = 3) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  facet_wrap(~measure) +
  scale_fill_manual(values = c("College (BA+)" = "#348ABD", "Non-College" = "#E24A33")) +
  labs(
    title = "Gender Gap in Labor Force Outcomes, Ages 20-29",
    subtitle = "Women minus Men. Negative = men have higher rate.",
    y = "Percentage Points", x = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

## Composition Check: Share with BA+ by Sex

How many 20-29 year-olds have a BA+ in each year? This matters because the college share has changed substantially, especially for women.

```{r composition}
comp <- df |>
  group_by(YEAR, MONTH, female) |>
  summarise(
    pct_college = 100 * sum(WTFINL[EDUC >= 111]) / sum(WTFINL),
    pct_in_school = 100 * sum(WTFINL[in_school]) / sum(WTFINL),
    .groups = "drop"
  ) |>
  group_by(YEAR, female) |>
  summarise(
    pct_college = mean(pct_college),
    pct_in_school = mean(pct_in_school),
    .groups = "drop"
  )

comp |>
  kable(digits = 1,
        col.names = c("Year", "Sex", "% College (BA+)", "% In School"),
        caption = "Composition of 20-29 Year-Olds")
```
